% ============================================================================
% Methods
% ============================================================================
\section{Methods}
\label{sec:methods}

Our aim in this project was to study how clusters in an artificial neural network model influence its dynamics. For clarity, we divide this goal into three main parts: model architecture, experimental setup, and statistical analysis.

The code was written using Python 3.13\footnote{https://www.python.org} and the artificial neural networks were implemented using Brian2\footnote{https://brian2.readthedocs.io} .


\subsection{Model}
\label{sec:model}

A good model architecture provides a framework, which should be biologically plausible, but at the same time simple enough to have as few tunable parameters as possible, computationally inexpensive and easy to grasp. This framework should provide us with a "sandbox" environment, facilitating an easy way to design and test hypotheses.

The uniform model was built with 4000 excitatory neurons and 1000 inhibitory neurons. The connection probability between neurons was $p^{EE} = 0.2$ for excitatory and $p^{EI}, p^{IE}, p^{II} = 0.5$ for all connections involving an inhibitory neuron. 
The strengths of the connections were $J^{EE} = 0.024, J^{IE} = 0.014, J^{EI} = -0.045$ and $J^{II} = -0.057$ ($J^{YX}$ denotes a connection from neuron group $X$ to neuron group $Y$.)


Individual neuronal dynamics were set with the following equations:
\begin{align}
\dot V &= \frac{\mu - V}{\tau_m} + I_{\text{syn}} \\
\dot I_{\text{syn}} &= \frac{x - I_{\text{syn}}}{\tau_2} \\
\dot x &= -\frac{x}{\tau_1}
\end{align}

If neuron \textbf{a} and \textbf{b} have a connection $\mathbf{a} \to \mathbf{b}$, then in the event of neuron \textbf{a} firing, $x_b = x_b + \frac{J^{BA}}{\tau_1}$.

The model used non-dimensional values in the simulations. 
The parameters were set as follows:
$\tau = 15ms$ for excitatory or $10ms$ for inhibitory neurons.
$\tau_1 = 1ms$ for both neuron types.
$\tau_2 = 3ms$ for excitatory or $2ms$ for inhibitory neurons.
Spike threshold $V_{th} = 1$
Reset potential $V_{re} = 0$
Refractory period $r = 5ms$
External input $\mu$ was sampled from a uniform distribution in the range $[1.1,1.2]$ for excitatory and $[1,1.05]$ for inhibitory neurons.
The initial $V$ was sampled from a uniform distribution in the range $[V_{re}, V_{th}]$

Dimensionality was introduced only when plotting the figures – the parameters were multiplied by $15mV$ and transformed so that $V_{th} = -50mV$ and $V_{re} = -65mV$.

Integration was performed using the Euler method, with timestep $dt = 0.1ms$

When constructing the clustered network, clusters were formed only between excitatory neurons. In a network with 4000 excitatory neurons, every cluster had 80 members. Clustering introduced two main changes to the network architecture: (1) the in-cluster connections were stronger (1.9x for our experiments); (2) Neurons had a higher chance of forming a synapse between a same-cluster counterpart.

The probability for forming excitatory connections inside clusters was set to $p_{in}^{EE}$, and to neurons in other clusters to $p_{out}^{EE}$. The probabilities were chosen such that the total average connection probability remained $p^{EE} = 0.2$. The ratio
\begin{equation}
\label{eq:ree_ratio}
R^{EE} = \frac{p_{in}^{EE}}{p_{out}^{EE}}
\end{equation}
determines the level of clustering in the network. In our experiments, we used $R^{EE} = 2.5$ in the clustered networks.


\subsection{Analysis}
\label{sec:analysis}

For understanding what kind of dynamics clustering introduces to neural networks, we ran parallel simulations on a uniformly connected network and a clustered network. The network was run for 3 seconds for all of our experiments. After simulations, the network behavior was analyzed using five key metrics: firing rate, Fano factor, correlation coefficients, autocorrelation, and cross-correlation. For the analysis, we discarded the first 1.5 seconds of the simulation and worked with the rest.

Firing rate for neuron i was calculated as
\begin{equation}
\label{eq:rate_definition}
r_i = \frac{N_i(t,t+\Delta t)}{\Delta t}
\end{equation}
  where $N_i(t,t+\Delta t)$ is the spike count of neuron \textbf{i} in a time interval $[t, t+\Delta t]$.

The Fano factor is a metric that captures spike variability. A Poisson process with a fixed underlying firing rate has a Fano factor of 1. Higher Fano factor values indicate a change in the neuron's underlying firing rate (for example, a transition from a low to a high firing rate state). The Fano factor for neuron \textbf{i} was calculated as 
\begin{equation}
\label{eq:fano_factor}
F_i =
\frac{\operatorname{Var}\big(N_i(t,t+\Delta t)\big)}
{\langle N_i(t,t+\Delta t) \rangle}
\end{equation}
The variance and mean of a neuron were calculated over 9 trials, using $\Delta t = 100ms$ and $t_{start} = 1.5s$

The correlation coefficients were calculated between neuron pairs. This metric tells us whether the spiking of two neurons is correlated or not. The correlation coefficient between neurons \textbf{i} and \textbf{j} was calculated as: 
\begin{equation}
\label{eq:correlation}
\rho_i =
\frac{\operatorname{Cov}\big(N_i(t,t+\Delta t), N_j(t,t+\Delta t)\big)}
{\sqrt{\operatorname{Var}\big(N_i(t,t+\Delta t)\big)
\operatorname{Var}\big(N_j(t,t+\Delta t)\big)}}
\end{equation}
Covariance and variance was calculated on overlapping windows within each trial and then averaged across the trials. In our experiments, we used $\Delta t = 50ms$ and $\Delta t = 100ms$ with a window step of $25ms$.

\textcolor{red}{part about autocorrelation and cross-correlation}
\textcolor{red}{part about fano factor vs change in $R_{EE}$}

Finally, we investigated how changing the input stimulus $\mu$ influences the network dynamics. For this, we raised the stimulus by a multiplier of 2x for the first cluster (80 neurons). The effect of elevated stimulus on the network was analyzed by calculating the Fano factor over 9 trials as before, but it was calculated for overlapping windows of $250ms$, with a shift of $30ms$, resulting in a time-dependent function for the Fano factor.
